Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-08-14 Report/Sections/models.tex

Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level (or coarse-scale) observations $Y \equiv \Big(Y_1, \ldots, Y_T\Big)$. The coarse-scale observations contain summaries of each curve such as curve length or maximum value. In addition, consider an associated sequence of within-curve (or fine-scale) observations $Y^* \equiv \Big(Y^*_1,\ldots,Y^*_T \Big)$, where $Y^*_{t} \equiv \Big(Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big)$ is a vector of fine-scale observations corresponding to curve $t$. The vector $Y^*_{t}$ is made up of function values from curve $t$, and we index each observation of $Y^*_t$ using $t^*$.

Hidden Markov models can describe state-switching Markovian processes in discrete time and will be the core structure of the models we use to describe $Y$ and $Y^*$. %Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain. However, when conditioned on the hidden state sequence observations are assumed to be independent. Therefore, classical HMMs do not hold when the observations exhibit other forms of significant correlation in time. Examples of this include continuous-time processes such as a Wiener process, systems which exhibit temporal dependencies acting at several different time scales \citep{Barajas:2017}, and periodic behaviour including the motion of an ideal spring.  
%\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things. You can use some of the text I removed from below}.
%To account for each of these common additional dependence structures,
We first describe the structure of a traditional HMM. We then describe a variation introduced by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either of these models can be used to model $Y$ or $Y^*_t$ for a particular curve. To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which use two distinct, but related, HMMs to jointly model these time series. Next, we describe how to reduce fine-scale correlation by transforming raw observations to meaningful summaries. In particular, we use the discrete Fourier transform (DFT) on a moving window over $Y^*$ to summarize its periodic behavior. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

@@

An HMM is comprised of a sequence of unobserved states $X = \big(X_1, \ldots, X_T\big)$ where $X_t$ is associated with the observation $Y_t$. Here we focus on the coarse-scale observations $Y$, but an HMM can also be used as a sub-model for a particular fine-scale process $Y_t^*$. The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
$\delta_i = \Pr\{X_1 = i\}.$
A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$. If $Y_t$ is a vector, we assume that the dimensions of the observations are independent $\left(f^{(i)}(\cdot) = \prod_{d} f^{(i)}_d(\cdot)\right)$, but this need not be the case in general.
%
Figure \ref{fig:models} represents the dependence structure of an HMM.

Using observed emissions, here denoted $y = (y_1,\ldots,y_T)$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$. We write the likelihood $\calL_{\text{HMM}}$ using the  well-known \textit{forward algorithm} \citep{Zucchini:2016}:
%
$$\calL_{\text{HMM}}(y;\Theta,\Gamma) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry  $f^{(i)}(y_t; \theta^{(i)})$.
%

Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows to sum to 1:
%
\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability. This simplifies likelihood maximization by removing constraints in the optimization problem. For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.

@@

The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence not accounted by the HMM state. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$ rather than only $X_t$. 
The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$.
%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
For example, one could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
Figure \ref{fig:models} shows a graphical representation of the dependence structure of a CarHMM.

The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation*}
    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation*}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.

@@

An HHMM accounts for different levels of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}. The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.   
%
In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$. The fine-scale state sequence in turn emits a sequence of fine-scale observations, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. This fine-scale process ($X_t^*, Y_t^*$) then makes up another HMM whose parameters depend on the value of $X_t$. Specifically, if $X_t=i$, then the distribution of $X_t^*$ is characterized by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary in general. The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$. Given the coarse-scale hidden states, $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.

Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM. For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.

Figure \ref{fig:models} represents the dependence structure for an HHMM. Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^* \equiv \left(y^*_1, \ldots,y^*_T\right)$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\[
\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
\Gamma^{*(i)}\right)$.

@@

In many applications where data is collected at high frequencies, intricate dependency structures may arise within the fine-scale process $Y^*$ which cannot be adequately modeled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behavior. To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behavior. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. 
%
%This moving window approach complicates the indexing of $Y_t^*$. Suppose a window size of $h$ observations is selected. Moving forward, we re-index $Y^*$ such that the fine-scale process associated with curve $t$ is denoted as $Y^*_t$ as before. However, the $t^{*th}$ \textit{window} of $Y^*_t$ is now denoted as $Y^*_{t,t^*}$, and the $t^{**th}$ \textit{observation} of $Y^*_{t,t^*}$ is denoted as $Y^*_{t,t^*,t^{**}}$. As a result, there are a total of $T^*_t$ windows associated with $Y_t^*$, each of which are of length $h$. Note that this relabeling may significantly reduce the length of $Y^*_t$ depending upon the window size $h$ and the step-size between windows.
%

For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem). We use $\Ztwo$ to capture the ``wiggliness" of the acceleration within a window. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other choices for window summary statistics are possible, including the dominant frequency and amplitude. Figure \ref{fig:models} represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM. To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$. The down sampling removes the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data, the method overall also reduces correlation in the raw observations that would due to the complex dependencies structure.

Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the HHMM structure. As such, the emission distributions and the likelihood remain identical to the original HHMM when transforming $Y^*$ to $\Z$. In order to clearly differentiate models, we refer to an HHMM with $\Z$ as fine-scale observations as an HHMM-DFT.

@@

By definition, an HHMM treats both the coarse-scale and the fine-scale processes as a traditional HMM. However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{HMM}}\left(y^*_t;\Theta^{*(i)}\right)$. 
On the fine scale, $Y^*_t$ can be modelled using \textit{any} parametric model which admits an easy-to-compute likelihood. The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
\[
\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.

>>>>

Consider a sequence of $T$ curves (or dives, in our context). Any particular dive $t$, $t = 1,
\ldots,T$, has an associated curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of within-curve (or fine-scale) observations $Y^*_{t} \equiv \Big(Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big)$. The vector $Y^*_{t}$ is made up of function values from curve $t$. We call the vectors of coarse-scale observations $Y \equiv \Big(Y_1, \ldots, Y_T\Big)$ and the collection of all fine-scale observations $Y^* \equiv \Big\{Y^*_1,\ldots,Y^*_T \Big\}$.

Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. %Traditional HMMs model process auto-correlation by assuming that the underlying behaviour of a processes follows a single Markov chain. However, when conditioned on the hidden state sequence observations are assumed to be independent. Therefore, classical HMMs do not hold when the observations exhibit other forms of significant correlation in time. Examples of this include continuous-time processes such as a Wiener process, systems which exhibit temporal dependencies acting at several different time scales \citep{Barajas:2017}, and periodic behaviour including the motion of an ideal spring.  
%\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things. You can use some of the text I removed from below}.
%To account for each of these common additional dependence structures,
We first detail the structure of a traditional HMM. We then review a variation introduced by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct, but related, HMMs that jointly model these time series. 
Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

@@

An HMM is comprised of a sequence of unobserved states $X = \big(X_1, \ldots, X_T\big)$ where $X_t$ is associated with the observation $Y_t$. Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model for the fine-scale sequence for a particular dive. The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
$\delta_i = \Pr(X_1 = i).$
A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. Namely, if $X_t = i$ then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
%
Figure \ref{fig:models}a represents the dependence structure of an HMM.

Using observed emissions, here denoted $y = (y_1,\ldots,y_T)$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$. We write the likelihood $\calL_{\text{HMM}}$ using the  well-known \textit{forward algorithm} \citep{Zucchini:2016}:
%
$$\calL_{\text{HMM}}(\Theta,\Gamma;y) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry  $f^{(i)}(y_t; \theta^{(i)})$.
%

Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows to sum to 1:
%
\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability. This simplifies likelihood maximization by removing constraints in the optimization problem. For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.

@@

The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. 
The first emission $Y_1$ is assumed to be fixed and known as an initial value which does not depend upon $X_1$.
%
We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
For example, one could assume that the conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\begin{align*}
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \text{ and } \\
\mathbb{V}(Y_t| Y_{t-1} = y_{t-1}, X_t = i) &= (\sigma^{(i)})^2.
\end{align*}
%
Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.

The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation*}
    \calL_{\text{CarHMM}}(\Theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation*}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.

@@

An HHMM accounts for different levels of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}. The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.   
%
In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$. The fine-scale state sequence in turn emits a sequence of fine-scale observations $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. For each $t$, the fine-scale process ($X_t^*, Y_t^*$) then follows another HMM whose parameters depend on the value of $X_t$. Specifically, if $X_t=i$, then the components of $X_t^*$ are a Markov chain with an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary in general. The distribution of $Y_{t,t^*}$ given $X_{t,t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$. Given the coarse-scale hidden states, the $T$ fine-scale HMMs $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.

Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM. For example, in the killer whale case study (section \ref{sec:data}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \text{i.e. } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$. Figure \ref{fig:models}c represents the dependence structure for an HHMM.

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\[
\calL_{\text{HHMM}}(\Theta,\Theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(\Theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$.

@@

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. 
%
%This moving window approach complicates the indexing of $Y_t^*$. Suppose a window size of $h$ observations is selected. Moving forward, we re-index $Y^*$ such that the fine-scale process associated with curve $t$ is denoted as $Y^*_t$ as before. However, the $t^{*th}$ \textit{window} of $Y^*_t$ is now denoted as $Y^*_{t,t^*}$, and the $t^{**th}$ \textit{observation} of $Y^*_{t,t^*}$ is denoted as $Y^*_{t,t^*,t^{**}}$. As a result, there are a total of $T^*_t$ windows associated with $Y_t^*$, each of which are of length $h$. Note that this relabeling may significantly reduce the length of $Y^*_t$ depending upon the window size $h$ and the step-size between windows.
%

For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*_t$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo$ corresponds to the ``wiggliness" of the acceleration data within a window. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude. To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$. The down-sampling avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the fine-scale HMM structure. As such, the emission distributions and the likelihood remain identical to the original HMM when transforming $Y^*$ to $\Z$. In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations as an HMM-DFT.

@@

By definition, an HHMM treats both the coarse-scale and the fine-scale processes as a traditional HMM. However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
\[
\calL_{\text{coarse}}(\Theta,\Theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{HMM}}\left(\Theta^{*(i)};y^*_t\right)$. 
On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods (or penalized likelihoods). The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(\Theta^{*(i)};y^*_t)$:
\[
\calL_{\text{coarse}}(\Theta,\Theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(\Theta^{*(i)};y^*_t\right)$.

Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here. 
For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. In particular, \citet{Xu:2018} model biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours. However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
Further, \citet{Bebbington:2007} and \citet{Borchers:2013} model data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, this model reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
In addition, \citep{Langrock:2018} use B-splines to model several aspects of an HMM, and this modified model admits a penalized likelihood which can be incorporated into the hierarchical structure here. 
%For these and countless other possible fine-scale models, added flexibility comes at a cost of increased model complexity.
